A weakly structured stem for human origins in Africa

### 2022-03-04 paper.tex
fix quotes

Archaeological sites from the Middle Stone Age (approx. 300kya-40kya) are widely
distributed across Africa, and are particularly well represented in the
northern, eastern and southern parts of the continent. Similarly, fossil crania
such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
anatomically derived \emph{Homo sapiens} features were also present across the
continent during this period. It has been difficult to reconcile these lines of
evidence with evidence from genomics, which have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population. It is unclear whether fossil specimens and archaeological sites
represent populations which contributed to our ancestors as population
precedents, or were local “dead-ends” from which contemporary \emph{Homo
sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of Homo sapiens beginning at least 300kya
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

To account for population structure prior to 135kya, three of our four models allowed for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or “merger”) events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
“multiple-merger” and the the “continuous-migration” models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the Western Africans and
Eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25kya.
The later Stem 2 contribution to the Western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
This may indicate that an ancestral Stem 2 population occupied Western or
Central Africa, broadly speaking. The differing proportions in the Nama and
Eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

By contrast, our inferred models paint a more consistent picture of the late
Middle Stone Age as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(Supporting Information).
During the Middle Stone Age, the Multiple Merger indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with a period of 
divergence, isolation and then a merger event (i.e. a “reticulation”)
out-performed models with bifurcating divergence and continuous gene flow.

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
modern groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later “Out of Africans”. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

>>>>

Archaeological sites from the Middle Stone Age (approx. 300kya-40kya) are widely
distributed across Africa, and are particularly well represented in the
northern, eastern and southern parts of the continent. Similarly, fossil crania
such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
anatomically derived \emph{Homo sapiens} features were also present across the
continent during this period. It has been difficult to reconcile these lines of
evidence with evidence from genomics, which have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population. It is unclear whether fossil specimens and archaeological sites
represent populations which contributed to our ancestors as population
precedents, or were local ``dead-ends'' from which contemporary \emph{Homo
sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of Homo sapiens beginning at least 300kya
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

To account for population structure prior to 135kya, three of our four models allowed for two or more
``stem'' populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
``multiple-merger'' and the the ``continuous-migration'' models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the Western Africans and
Eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25kya.
The later Stem 2 contribution to the Western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
This may indicate that an ancestral Stem 2 population occupied Western or
Central Africa, broadly speaking. The differing proportions in the Nama and
Eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

By contrast, our inferred models paint a more consistent picture of the late
Middle Stone Age as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(Supporting Information).
During the Middle Stone Age, the Multiple Merger indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with a period of 
divergence, isolation and then a merger event (i.e. a ``reticulation'')
out-performed models with bifurcating divergence and continuous gene flow.

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
modern groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.
